Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-04 paper.Rmd
replace "BS" with "B&S"

@rost:etal:2009,  GLOBUS, 
# Review of Blair and Sambanis 2020

@blair2020forecasting (BS for the sake of brevity hereafter) aim to examine whether theory adds value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed. They find that it does. They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,000 covariates.

BS specifically examine three questions:

@@

To assess the first two questions BS use global datasets covering all major countries from 2001 to 2015. Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years. The main outcome is civil war onset, measured using Sambanis' civil war dataset.

@@

The corresponding results for each question are shown in BS Tables 1 and 2, which we will replicate further below.

To examine the first question, BS compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

@@

The corresponding results are shown in BS Table 2.

Finally, BS used the Escalation model to create forcasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed. This is summarized in BS Table 3.

@@

While attempting to replicate BS's results, we found several issues.

@@

The most consequential issue that we found is that all AUC-ROC values reported in BS Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves. A reference to smoothing is made in a single sentence in the paper (p. 12):

@@

Figure 1 shows our replication of both the smoothed ROC curves BS report, and the actual ROC curves on the right.

![Replication of BS Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)

ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data. In this case there are also groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached. For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in BS Table 1, there are only 11 civil war onset cases in the test set. Thus the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates. We can only speculate, but maybe the very "step"-like nature of the ROC curves plus the presence of unusual diagonal lines accounts for the decision to use smoothed ROC curves.

As it turns out, this decision has a dramatic impact on the AUC-ROC calculations. As we will show in the results further below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and differentially impacts the models that are being compared. In fact, BS's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC.

To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves. We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations. It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the BS results, it minimally would have warranted an explicit discussion in the paper. This is not the case, and it is only mentioned in the sentence we quote above.

@@

The "Weighted by PITF" model is described as follows in BS, page 19:

@@

Similarly, the "PITF Split Population" model appears to be incorrectly implemented. BS describe it as, on page 20:

@@

The intention clearly was to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they are both run on the exact same training data, which consists of the full training data all other models are run on. The models are also identical otherwise, i.e. they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the BS replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same RNG seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]

@@

Lastly, the AUC-ROC values reported in the original BS Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in Table 2. Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in BS Table 2 are actually not comparable to the other, and any conclusions drawn from such comparison are potentially incorrect.

We fix this issue by only using predictions for common joint subset of cases that all models have non-missing predictions for. The original BS Table 2 1-month have *N*=11,806--12,495 versus a common joint subset of 9,811, and for the 6-month row *N*=2,070--2,233 versus *N*=1,915 for the common joint subset.

@@

BS show a confusiong matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war onset, in the replication code they are actually scored using the much more common incidence of civil war, i.e. including ongoing civil wars as "1"'s.

@@

Table \ref{tab:table1-smooth} is our replication of BS Table 1 with smoothed AUC-ROC. The results differ slightly from the original BS Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models. It is the case that BS set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel. In any case, the interpretation of results should not be sensitive to random variation, i.e. it should not depend on using a specific RNG seed. On the basis of these results, BS conclude that the Escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.

@@

table1_smooth %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of BS Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
    label = "table1-smooth") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
```

Table \ref{tab:table1-nosmooth} shows a version of BS Table 1 with the conventional non-smoothed ROC curves. The Average model outperforms the Escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus the original BS conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.

@@

table1_nosmooth %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of BS Table 1 \\textit{without} smoothed ROC curves; test set AUC-ROC for various models",
    label = "table1-nosmooth") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
```

@@

Table \ref{tab:table2-fixed} shows our replication of BS Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for. Table \ref{tab:table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.

@@

table2_nosmooth %>%
  dplyr::select(-Model, -horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of BS Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution",
    label = "table2-fixed") %>%
  column_spec(2:6, width = "2cm") %>%
  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
  footnote("Differences from the original BS Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
```

BS interpret the results as follows, on page 20:^[We list the "Overall, ..." interpretation out of order, last, for clarity.]

@@

```{r}
table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
table1_benefit %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Smoothing advantage for BS Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves",
    label = "table1-benefit") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))

table2_benefit <- read_csv("data/table2-smooth-benefit.csv") 
table2_benefit %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Smoothing advantage for BS Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves",
    label = "table2-benefit") %>% 
  column_spec(2:6, width = "2cm") %>%
  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1))
```

What impact did the ROC smoothing have overall on the results reported in BS Tables 1 and 2? Tables \ref{tab:table1-benefit} and \ref{tab:table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC. Positive values indicate that smoothing increased a model's apparent performance. The Escalation model is the only model that consistently had a benefit from smoothing. For all 8 other models, smoothing sometimes gave a benefit, sometimes not.

A more dramatic difference stands out when we consider the overall average impact of smoothing across all flavors of a model reported in the columns of the tables. The Average, Goldstein, Quad, PITF Split Population, and Weighted by PITF models are slightly hurt by smoothing, but on the order of less than 0.01 in absolute magnitude. The "With PITF Predictors" model is hurt on the order of 0.02, i.e. it appears substantially worse with smoothing. The CAMEO and PITF Only models benefit slightly, on the order of 0.01 or less. The Escalation Only model on the other hand has an average boost of 0.058 to it's AUC-ROC from smoothing. Considering the spread of AUC-ROC values if we compare across rows in BS Tables 1 and 2, that boost is substantial.

@@

What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.

|Hyperparameter|Default heuristic|Default values (Escalation)|B&S value|
|--------------|-----------------|--------------------------|---------|
|type          |                 | classification           | regression|
|ntree         |                 | 500                      | 100,000 or 1e6 |
|mtry          |`floor(sqrt(ncol(x)))`|3|3|
|replace       |                 | true | false |
|sampsize | `nrow(x)` if replace, | 11,869 | 100 or 500 |
||else `ceiling(.632*nrow(x))`|||
|nodesize | 1 for classification | 1 | 1 |
|maxnodes | | null | 5 or 10 |

There is a stark contrast in the default RF settings and the way Blair and Sambanis use the RF models. The default approach is to train a relatively small number (ntree; 500) of classification trees, but where each tree is fairly big in that it is trained on data that has the same number of rows as the training data, albeit sampled with replacement (replace is true; sampsize is 11,869), and allowed to grow fairly deep (this is governed by nodesize (1), which is the minimum size a terminal node must have). In contrast, Blair and Sambanis grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).

This approach only works due to the other unorthodox choice, which is to use regression, not classification, trees. Trying to use classification trees with the other parameter settings in fact does not work at all because it is almost guaranteed that a sample of 100 from the 11,869 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample. As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:

@@

The RF approach BS use is essentially similar to the probability forests proposed in Malley 2012. It is not clear if this is intentional or not, as the paper is not cited.

We tried to create versions of the RF models that use the regular framework for classification forests, but were unable to create models that approach the BS RFs test set performance.

How were these hyper-parameter values determined? Do they work better for some models than others?

@@

BS has a clear misunderstanding of the role that out-of-sample prediction can plan in analysis.  On the one hand it can be used for simple forecasting while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.  They further misrepresent cross-validation which is not about fishing for results, as claimed by BS.  The current state-of-the-art uses prediction and cross-validation most frequently to provide supportive evidence that is independent of the estimation procedure and the in-sample data. It is incorrect to paint this procedure as atheoretical since many studies will have some explanation for how the model was constructed. These procedures may simply be used to provide evidence for a theoretical argument.  Though that argument may be unconvincing and/or atheoretical.

The BS approach that is advocated is to use theory to guide prediction. But theory is a quite ambiguous and undefined concept.  It is not a procedure.  What they actually do is to create a model with four right hand side variables that is supposed to capture a complicated repression-dissent dynamic.  The dynamic is probably not linear, but their model is. There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.  As such their baseline comparison is hardly a standard bearer for strong theory.

@@

We encountered several issues in the code underlying the BS analysis.

@@

When we fix these issue and perform an updated analysis, the conclusions BS draw all essentially overturn. In other words, BS findings are based on a faulty analysis, and invalid.

Using the same analysis BS intend to use, we in fact find that:

@@

**AB: add a table comparing BS original claims and updated results.**

@@

table2_full %>%
  select(horizon, column, N_test, cases) %>%
  pivot_wider(names_from = column, values_from = N_test) %>%
  arrange(horizon, desc(cases)) %>%
  rename(Cases = cases, Horizon = horizon) %>%
  select(Cases, Horizon, everything()) %>%
  arrange(desc(Cases), Horizon) %>%
  knitr::kable(
    "latex", booktabs = TRUE, digits = 0,
    caption = "Number of valid test predictions for each cell in BS Table 2",
    label = "table2-N") %>%
  column_spec(3:7, width = "2cm") %>%
  collapse_rows(1:2, row_group_label_position = "stack")

table2_full %>%
  pivot_longer(auc_roc:auc_roc_smoothed, names_to = "Smoothed ROC",
               values_to = "auc_roc") %>%
  mutate(`Smoothed ROC` = case_when(
    `Smoothed ROC`=="auc_roc" ~ "No",
    TRUE ~ "Yes")
  ) %>%
  select(horizon, column, cases, `Smoothed ROC`, auc_roc) %>%
  pivot_wider(names_from = column, values_from = auc_roc) %>%
  rename(Cases = cases, Horizon = horizon) %>%
  select(Cases, Horizon, `Smoothed ROC`, everything()) %>%
  arrange(desc(Cases), Horizon, desc(`Smoothed ROC`)) %>%
  knitr::kable(
    "latex", booktabs = TRUE, digits = 2,
               caption = "Replication of BS Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N",
    label = "table2-full") %>%
  column_spec(3, width = "1.5cm") %>%
  column_spec(4:8, width = "2cm") %>%
  collapse_rows(1:3, row_group_label_position = "stack")

>>>>

@rost:etal:2009,  GLOBUS,

# Review of Blair and Sambanis 2020

@blair2020forecasting (B&S for the sake of brevity hereafter) aim to examine whether theory adds value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed. They find that it does. They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,000 covariates.

B&S specifically examine three questions:

@@

To assess the first two questions B&S use global datasets covering all major countries from 2001 to 2015. Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years. The main outcome is civil war onset, measured using Sambanis' civil war dataset.

@@

The corresponding results for each question are shown in B&S Tables 1 and 2, which we will replicate further below.

To examine the first question, B&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

@@

The corresponding results are shown in B&S Table 2.

Finally, B&S used the Escalation model to create forcasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed. This is summarized in B&S Table 3.

@@

While attempting to replicate B&S's results, we found several issues.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves. A reference to smoothing is made in a single sentence in the paper (p. 12):

@@

Figure 1 shows our replication of both the smoothed ROC curves B&S report, and the actual ROC curves on the right.

![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)

ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data. In this case there are also groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached. For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates. We can only speculate, but maybe the very "step"-like nature of the ROC curves plus the presence of unusual diagonal lines accounts for the decision to use smoothed ROC curves.

As it turns out, this decision has a dramatic impact on the AUC-ROC calculations. As we will show in the results further below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and differentially impacts the models that are being compared. In fact, B&S's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC.

To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves. We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations. It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper. This is not the case, and it is only mentioned in the sentence we quote above.

@@

The "Weighted by PITF" model is described as follows in B&S, page 19:

@@

Similarly, the "PITF Split Population" model appears to be incorrectly implemented. B&S describe it as, on page 20:

@@

The intention clearly was to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they are both run on the exact same training data, which consists of the full training data all other models are run on. The models are also identical otherwise, i.e. they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same RNG seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]

@@

Lastly, the AUC-ROC values reported in the original B&S Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in Table 2. Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B&S Table 2 are actually not comparable to the other, and any conclusions drawn from such comparison are potentially incorrect.

We fix this issue by only using predictions for common joint subset of cases that all models have non-missing predictions for. The original B&S Table 2 1-month have *N*=11,806--12,495 versus a common joint subset of 9,811, and for the 6-month row *N*=2,070--2,233 versus *N*=1,915 for the common joint subset.

@@

B&S show a confusiong matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war onset, in the replication code they are actually scored using the much more common incidence of civil war, i.e. including ongoing civil wars as "1"'s.

@@

Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC. The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models. It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel. In any case, the interpretation of results should not be sensitive to random variation, i.e. it should not depend on using a specific RNG seed. On the basis of these results, B&S conclude that the Escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.

@@

table1_smooth %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of B\\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
    label = "table1-smooth") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
```

Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves. The Average model outperforms the Escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus the original B&S conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.

@@

table1_nosmooth %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of B\\&S Table 1 \\textit{without} smoothed ROC curves; test set AUC-ROC for various models",
    label = "table1-nosmooth") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
```

@@

Table \ref{tab:table2-fixed} shows our replication of B&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for. Table \ref{tab:table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.

@@

table2_nosmooth %>%
  dplyr::select(-Model, -horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Replication of B\\&S Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution",
    label = "table2-fixed") %>%
  column_spec(2:6, width = "2cm") %>%
  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
  footnote("Differences from the original B\\&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
```

B&S interpret the results as follows, on page 20:^[We list the "Overall, ..." interpretation out of order, last, for clarity.]

@@

```{r}
table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
table1_benefit %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Smoothing advantage for B\\&S Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves",
    label = "table1-benefit") %>% 
  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))

table2_benefit <- read_csv("data/table2-smooth-benefit.csv") 
table2_benefit %>%
  select(-horizon) %>%
  knitr::kable(
    digits = 2, booktabs = TRUE, 
    caption = "Smoothing advantage for B\\&S Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves",
    label = "table2-benefit") %>% 
  column_spec(2:6, width = "2cm") %>%
  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1))
```

What impact did the ROC smoothing have overall on the results reported in B&S Tables 1 and 2? Tables \ref{tab:table1-benefit} and \ref{tab:table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC. Positive values indicate that smoothing increased a model's apparent performance. The Escalation model is the only model that consistently had a benefit from smoothing. For all 8 other models, smoothing sometimes gave a benefit, sometimes not.

A more dramatic difference stands out when we consider the overall average impact of smoothing across all flavors of a model reported in the columns of the tables. The Average, Goldstein, Quad, PITF Split Population, and Weighted by PITF models are slightly hurt by smoothing, but on the order of less than 0.01 in absolute magnitude. The "With PITF Predictors" model is hurt on the order of 0.02, i.e. it appears substantially worse with smoothing. The CAMEO and PITF Only models benefit slightly, on the order of 0.01 or less. The Escalation Only model on the other hand has an average boost of 0.058 to it's AUC-ROC from smoothing. Considering the spread of AUC-ROC values if we compare across rows in B&S Tables 1 and 2, that boost is substantial.

@@

What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated. Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most the models reported in the paper.

As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly. The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictived probabilities by averaging over the "0" or "1" votes from all constituent decision trees. The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number. From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data. These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.

\begin{table}
\caption{\label{tab:hp} Random forest (\texttt{randomForest()}) default versus B\&S hyperparameters}
\begin{tabular}{l>{\raggedright\arraybackslash}p{2in}ll}
\toprule
Hyperparameter & Default heuristic & Default values (Escalation) & B\&S value \\
\midrule
type & & classification & regression \\
ntree & & 500 & 100,000 or 1e6 \\
mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3 \\
replace & & true & false \\
sampsize & \texttt{nrow(x)} if replace, else \texttt{ceiling(.632*nrow(x))} & 11,869 & 100 or
500 \\
nodesize & 1 for classification & 1 & 1 \\
maxnodes & & null & 5 or 10 \\
\bottomrule
\end{tabular}
\end{table}

B&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement,  of the available training data. This approach only works due to the choice to use regression, not classification, trees. Trying to use classification trees with the other parameter settings not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample. As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:

@@

As it turns out, using regression random forests for this kind of binary classification problem in order to obtain probability estimates matches the probability random forest approach suggested and positively evaluated in @malley:etal:2012, and which is used in another prominent R implementation of random forests.^[The **ranger** package.] It is not clear whether this is intentional, as the Malley paper is not cited in B&S.

In any case, B&S's random forest approach appears to work really well. We tried to construct classification random forests tuned via cross-validation on the training data set partition, i.e. without touching the test data, but were unable to develop models that consistently match the B&S random forest method in both cross-validated out-of-sample training predictions and test set predictions.

Given that they are relatively unorthodox, yet appear to work very well, we wonder how the hyper-parameter values were determined. Two specific concerns are that this was not done with an eye towards test set accuracy, which would invalidate the independence of the out-of-sample test set, and whether the specific hyper-parameter values are optimized for only one model, or were optimized and found to work well for all models. There is no discussion of the random forest tuning strategy or how the specific hyper-parameter methods were determined in the paper.

@@

B&S has a clear misunderstanding of the role that out-of-sample prediction can plan in analysis.  On the one hand it can be used for simple forecasting while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.  They further misrepresent cross-validation which is not about fishing for results, as claimed by B&S.  The current state-of-the-art uses prediction and cross-validation most frequently to provide supportive evidence that is independent of the estimation procedure and the in-sample data. It is incorrect to paint this procedure as atheoretical since many studies will have some explanation for how the model was constructed. These procedures may simply be used to provide evidence for a theoretical argument.  Though that argument may be unconvincing and/or atheoretical.

The B&S approach that is advocated is to use theory to guide prediction. But theory is a quite ambiguous and undefined concept.  It is not a procedure.  What they actually do is to create a model with four right hand side variables that is supposed to capture a complicated repression-dissent dynamic.  The dynamic is probably not linear, but their model is. There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.  As such their baseline comparison is hardly a standard bearer for strong theory.

@@

We encountered several issues in the code underlying the B&S analysis.

@@

When we fix these issue and perform an updated analysis, the conclusions B&S draw all essentially overturn. In other words, B&S findings are based on a faulty analysis, and invalid.

Using the same analysis B&S intend to use, we in fact find that:

@@

**AB: add a table comparing B&S original claims and updated results.**

@@

table2_full %>%
  select(horizon, column, N_test, cases) %>%
  pivot_wider(names_from = column, values_from = N_test) %>%
  arrange(horizon, desc(cases)) %>%
  rename(Cases = cases, Horizon = horizon) %>%
  select(Cases, Horizon, everything()) %>%
  arrange(desc(Cases), Horizon) %>%
  knitr::kable(
    "latex", booktabs = TRUE, digits = 0,
    caption = "Number of valid test predictions for each cell in B\\&S Table 2",
    label = "table2-N") %>%
  column_spec(3:7, width = "2cm") %>%
  collapse_rows(1:2, row_group_label_position = "stack")

table2_full %>%
  pivot_longer(auc_roc:auc_roc_smoothed, names_to = "Smoothed ROC",
               values_to = "auc_roc") %>%
  mutate(`Smoothed ROC` = case_when(
    `Smoothed ROC`=="auc_roc" ~ "No",
    TRUE ~ "Yes")
  ) %>%
  select(horizon, column, cases, `Smoothed ROC`, auc_roc) %>%
  pivot_wider(names_from = column, values_from = auc_roc) %>%
  rename(Cases = cases, Horizon = horizon) %>%
  select(Cases, Horizon, `Smoothed ROC`, everything()) %>%
  arrange(desc(Cases), Horizon, desc(`Smoothed ROC`)) %>%
  knitr::kable(
    "latex", booktabs = TRUE, digits = 2,
               caption = "Replication of B\\&S Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N",
    label = "table2-full") %>%
  column_spec(3, width = "1.5cm") %>%
  column_spec(4:8, width = "2cm") %>%
  collapse_rows(1:3, row_group_label_position = "stack")
